An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-09-10 Project_0/manuscript/manuscript_R1.tex
saving work on revision

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{08 August, 2025}

@@

\hypertarget{abstract}{%
\subsection*{Abstract}\label{abstract}}
\addcontentsline{toc}{subsection}{Abstract}

@@

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

@@

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
whether convergence and Akaike Information Criterion (AIC) can correctly
determine the source of process error and the existence of a SRR. We
also evaluate when retrospective patterns occur and the degree of bias
in the outputs of the assessment model that are important for
management.

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

@@

\hypertarget{operating-models}{%
\subsection*{Operating models}\label{operating-models}}
\addcontentsline{toc}{subsection}{Operating models}

\hypertarget{population}{%
\subsubsection*{Population}\label{population}}
\addcontentsline{toc}{subsubsection}{Population}

@@

\hypertarget{fleets}{%
\subsubsection*{Fleets}\label{fleets}}
\addcontentsline{toc}{subsubsection}{Fleets}

@@

\hypertarget{indices}{%
\subsubsection*{Indices}\label{indices}}
\addcontentsline{toc}{subsubsection}{Indices}

@@

\hypertarget{observation-uncertainty}{%
\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}}
\addcontentsline{toc}{subsubsection}{Observation Uncertainty}

@@

\hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
\subsubsection*{Operating models with random effects on numbers at
age}\label{operating-models-with-random-effects-on-numbers-at-age}}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on numbers at age}

@@

\hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
\subsubsection*{Operating models with random effects on natural
mortality}\label{operating-models-with-random-effects-on-natural-mortality}}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on natural mortality}

@@

\hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
\subsubsection*{Operating models with random effects on fleet
selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on fleet selectivity}

@@

\hypertarget{operating-models-with-random-effects-on-index-catchability}{%
\subsubsection*{Operating models with random effects on index
catchability}\label{operating-models-with-random-effects-on-index-catchability}}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on index catchability}

@@

\hypertarget{estimation-models}{%
\subsection*{Estimation models}\label{estimation-models}}
\addcontentsline{toc}{subsection}{Estimation models}

@@

\hypertarget{measures-of-reliability}{%
\subsection*{Measures of reliability}\label{measures-of-reliability}}
\addcontentsline{toc}{subsection}{Measures of reliability}

\hypertarget{convergence}{%
\subsubsection*{Convergence}\label{convergence}}
\addcontentsline{toc}{subsubsection}{Convergence}

@@

\hypertarget{aic-for-model-selection}{%
\subsubsection*{AIC for model selection}\label{aic-for-model-selection}}
\addcontentsline{toc}{subsubsection}{AIC for model selection}

@@

\hypertarget{bias}{%
\subsubsection*{Bias}\label{bias}}
\addcontentsline{toc}{subsubsection}{Bias}

For a given model attribute we calculated the relative error \[
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] from fitting a given estimating model to simulated data set \(i\)
configured for a given OM where \(\widehat \theta_i\) and \(\theta_i\)
are the estimated and true values for simulation \(i\). We estimated
bias as the median of the relative errors across all simulations for a
given OM and EM combination. We constructed 95\% confidence intervals
for the median relative bias using the binomial distribution approach as
in Miller and Hyun (2018) and Stock and Miller (2021). We present
results for bias in terminal year estimates of SSB and recruitment,
Beverton-Holt stock recruit parameters (\(a\) and \(b\)), and median
natural mortality rate. Results for terminal year fishing mortality were
strongly negatively correlated with those for SSB and are provided in
the Supplementary Materials.

\hypertarget{mohns-rho}{%
\subsubsection*{\texorpdfstring{Mohn's
\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho}}
\addcontentsline{toc}{subsubsection}{Mohn's \(\rho\)}

We calculated Mohn's \(\rho\) for SSB, fully-selected fishing mortality,
and recruitment for each EM (Mohn 1999). We fit 7 peels for each EM and
calculated median 95\% confidence intervals for Mohn's \(\rho\) using
the same methods as that for relative bias.

\hypertarget{summarizing-results-across-oms}{%
\subsubsection*{Summarizing results across
OMs}\label{summarizing-results-across-oms}}
\addcontentsline{toc}{subsubsection}{Summarizing results across OMs}

Because of the numerous factors investigated for each of the ways we
assessed reliability of the state-space assessment models, we used two
methods to summarize the most important factors that explain differences
for each measure of reliability. The first method was fitting regression
models with the response being each of the measures of reliability
described above and predictor variables were defined based on OM and EM
characteristics. For convergence results we performed logistic
regressions and for AIC results we performed multinomial regressions for
the type of process error that provided the lowest AIC. For other
measures of reliability we fit standard linear regression models.
Because relative bias and Mohn's \(\rho\) for the various parameters are
bounded below at -1, we used as a response the log of the absolute
log-scale errors \begin{equation}\label{bias_regression_response}
y_i = \log\left|\log\left(\widehat \theta_i\right) - \log(\left(\theta_i\right)\right|,
\end{equation} so that lower values measure less bias whatever the sign.
For relative bias we also restricted the fits to simulations where the
estimated CV of the parameter or model output was less than 100\%
because values greater than this would typically lead the assessment
scientist away from the particular configuration of the estimating
model. For all regressions we fit models with factors included
individually and with all factors and with all second and third order
interactions. For the multinomial regression, we used the \verb|vglm|
function from the VGAM package (Yee 2008; Yee 2015). We tabulated
percent reduction in residual deviance for each of regression fits.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the primary factors that split the
values for each measure of reliability and subsets of the measures that
were alloted to each branch of the splits. Classification trees were fit
to categorical measures (convergence and AIC) and regression trees were
fit to the other measures with continuous scales. The response variables
were the same as the regressions for the deviance reduction analyses. We
used the \verb|rpart| function in the rpart package
(\url{https://cran.r-project.org/package=rpart}) to fit trees and we
pruned them to show only the primary branches for clarity. Generally,
the factors defining primary splits for the trees corresponded to the
factors with the largest reductions in deviance fitting the regression
models.

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

\hypertarget{convergence-performance}{%
\subsection*{Convergence performance}\label{convergence-performance}}
\addcontentsline{toc}{subsection}{Convergence performance}

\hypertarget{deviance-reduction-and-classification-trees}{%
\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\hypertarget{full-results}{%
\subsubsection*{Full results}\label{full-results}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\hypertarget{aic-performance-for-process-error-structure}{%
\subsection*{AIC performance for process error
structure}\label{aic-performance-for-process-error-structure}}
\addcontentsline{toc}{subsection}{AIC performance for process error
structure}

\hypertarget{deviance-reduction-and-classification-trees-1}{%
\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees-1}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\hypertarget{full-results-1}{%
\subsubsection*{Full results}\label{full-results-1}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\hypertarget{aic-performance-for-the-stock-recruit-relationship}{%
\subsection*{AIC performance for the stock-recruit
relationship}\label{aic-performance-for-the-stock-recruit-relationship}}
\addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
relationship}

\hypertarget{deviance-reduction-and-classification-trees-2}{%
\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees-2}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\hypertarget{full-results-2}{%
\subsubsection*{Full results}\label{full-results-2}}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\hypertarget{bias-1}{%
\subsection*{Bias}\label{bias-1}}
\addcontentsline{toc}{subsection}{Bias}

\hypertarget{deviance-reduction-and-regression-trees}{%
\subsubsection*{Deviance reduction and regression
trees}\label{deviance-reduction-and-regression-trees}}
\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
trees}

We observed different effects of factors on deviance reduction for
errors in estimation of the two Beverton-Holt stock-recruit parameters.
When analyses were For bias of the density-dependent parameter \(b\) as
measured by the log absolute value of log-scale errors, deviance was not
reduced by any of the individual OM or EM factors except for variation
in recruitment \(\sigma_R\) for R and R+S OMs. However, second and(or)
third order interactions of the OM and EM factors also provided large
reductions in deviance for all OM process error types which indicated
that some combinations of factors could reduce bias for the \(b\)
parameter (Table \ref{}). Despite the explanation variation in the \(b\)
estimation erorrs, bias was still significant regardless of the
combination of OM and EM factors (Figures regtrees and full results).

\hypertarget{full-results-3}{%
\subsubsection*{Full results}\label{full-results-3}}
\addcontentsline{toc}{subsubsection}{Full results}

\hypertarget{spawning-stock-biomass}{%
\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}}
\addcontentsline{toc}{paragraph}{Spawning stock biomass}

@@

\hypertarget{beverton-holt-parameters}{%
\subsubsection*{Beverton-Holt
parameters}\label{beverton-holt-parameters}}
\addcontentsline{toc}{subsubsection}{Beverton-Holt parameters}

@@

\hypertarget{median-natural-mortality-rate}{%
\subsubsection*{Median natural mortality
rate}\label{median-natural-mortality-rate}}
\addcontentsline{toc}{subsubsection}{Median natural mortality rate}

@@

\hypertarget{discussion}{%
\section*{Discussion}\label{discussion}}
\addcontentsline{toc}{section}{Discussion}

\hypertarget{convergence-1}{%
\subsection*{Convergence}\label{convergence-1}}
\addcontentsline{toc}{subsection}{Convergence}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{10 September, 2025}

@@

\subsection*{Abstract}\label{abstract}
\addcontentsline{toc}{subsection}{Abstract}

@@

\section*{Introduction}\label{introduction}
\addcontentsline{toc}{section}{Introduction}

@@

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
whether convergence and Akaike Information Criterion (AIC) can correctly
determine the source of process error and the existence of a SRR. We
also evaluate when retrospective patterns occur and the degree of bias
in outputs of the assessment model that are important for management.

\section*{Methods}\label{methods}
\addcontentsline{toc}{section}{Methods}

@@

\subsection*{Operating models}\label{operating-models}
\addcontentsline{toc}{subsection}{Operating models}

\subsubsection*{Population}\label{population}
\addcontentsline{toc}{subsubsection}{Population}

@@

\subsubsection*{Fleets}\label{fleets}
\addcontentsline{toc}{subsubsection}{Fleets}

@@

\subsubsection*{Indices}\label{indices}
\addcontentsline{toc}{subsubsection}{Indices}

@@

\subsubsection*{Observation Uncertainty}\label{observation-uncertainty}
\addcontentsline{toc}{subsubsection}{Observation Uncertainty}

@@

\subsubsection*{Operating models with random effects on numbers at
age}\label{operating-models-with-random-effects-on-numbers-at-age}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on numbers at age}

@@

\subsubsection*{Operating models with random effects on natural
mortality}\label{operating-models-with-random-effects-on-natural-mortality}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on natural mortality}

@@

\subsubsection*{Operating models with random effects on fleet
selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on fleet selectivity}

@@

\subsubsection*{Operating models with random effects on index
catchability}\label{operating-models-with-random-effects-on-index-catchability}
\addcontentsline{toc}{subsubsection}{Operating models with random
effects on index catchability}

@@

\subsection*{Estimation models}\label{estimation-models}
\addcontentsline{toc}{subsection}{Estimation models}

@@

\subsection*{Measures of reliability}\label{measures-of-reliability}
\addcontentsline{toc}{subsection}{Measures of reliability}

\subsubsection*{Convergence}\label{convergence}
\addcontentsline{toc}{subsubsection}{Convergence}

@@

\subsubsection*{AIC for model selection}\label{aic-for-model-selection}
\addcontentsline{toc}{subsubsection}{AIC for model selection}

@@

\subsubsection*{Bias}\label{bias}
\addcontentsline{toc}{subsubsection}{Bias}

For a given model attribute we calculated the relative error
\begin{equation}\label{relerror}
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\end{equation} from fitting a given estimating model to simulated data
set \(i\) configured for a given OM where \(\widehat \theta_i\) and
\(\theta_i\) are the estimated and true values for simulation \(i\). We
estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. We constructed 95\%
confidence intervals for the median relative bias using the binomial
distribution approach as in Miller and Hyun (2018) and Stock and Miller
(2021). We present results for bias in terminal year estimates of SSB
and recruitment, Beverton-Holt stock recruit parameters (\(a\) and
\(b\)), and median natural mortality rate. Results for terminal year
fishing mortality were strongly negatively correlated with those for SSB
and are provided in the Supplementary Materials.

\subsubsection*{\texorpdfstring{Mohn's
\(\rho\)}{Mohn's \textbackslash rho}}\label{mohns-rho}
\addcontentsline{toc}{subsubsection}{Mohn's \(\rho\)}

We calculated Mohn's \(\rho\) for SSB, fishing mortality (averaged over
all age classes), and recruitment for each EM (Mohn 1999). We fit
\(P =7\) peels to each simulated data set and calculated Mohn's \(\rho\)
for a given attribute \(\theta\) as \begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
attribute \(\theta\) in year \(i\) from a model fit using data up to
year \(j\). For each EM we calculated median and 95\% confidence
intervals using the same methods as that for relative bias.

\subsubsection*{Summarizing results across OM and EM
attributes}\label{summarizing-results-across-om-and-em-attributes}
\addcontentsline{toc}{subsubsection}{Summarizing results across OM and
EM attributes}

The measures of central tendency and variability of of observed values
for specific OM and EM attributes (e.g., low or high observation error)
that we described above can indicate scenarios that provide better or
poorer reliability. However, the OM and EM attributes that we
investigated are numerous, so we used two methods to summarize the most
important factors for differences in results. The first method was
fitting regression models with the response being each of the measures
of reliability described above and predictor variables were defined
based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
al. 2017; Harwell et al. 2018). For the binary results for convergence
and AIC-based selection of a stock-recruit relationship, we performed
logistic regressions. For results for AIC-based selection of EM process
error type (multiple categories) we performed multinomial regressions.
For other measures of reliability we fit standard linear regression
models. Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\)
for the various parameters are bounded below at -1, we used a
transformation of these values
\begin{equation}\label{bias_regression_response}
y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
log-scale error. We omitted simulation estimates equal to zero (RE =
-1). For all regressions we fit separate models with individual factors
included, with all factors combined, with including all second order
interactions, and including all third order interactions. For the
multinomial regression, we used the \verb|vglm| function from the VGAM
package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
deviance for each of regression fits. We did not perform formal
statistical analyses of effects of OM and EM attributes on results
(e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the OM and EM attributes, and their
interactions, split the values for each measure of reliability (e.g.,
Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
for categorical measures (convergence and AIC) and regression trees for
the other measures with continuous scales (relative error and Mohn's
\(\rho\)). The response variables were the same as the regressions for
the deviance reduction analyses. We used the \verb|rpart| function in
the rpart package (\url{https://cran.r-project.org/package=rpart}) to
fit trees and, for clarity we pruned them to show the primary branches.

\section*{Results}\label{results}
\addcontentsline{toc}{section}{Results}

\subsection*{Convergence performance}\label{convergence-performance}
\addcontentsline{toc}{subsection}{Convergence performance}

\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

EM process error assumption was the single attribute that resulted in
the largest percent reduction in deviance for all OM process error types
other than R+S OMs where the EM M assumption explained the most residual
deviance (Table \ref{convergence_PRD_table}). However, also including
interactions of OM and EM factors also provided large reductions in
residual deviance, suggesting successful convergence depended on a
combination OM and EM attributes. Classification trees for each OM
process error type, all had the primary branch defined using the same
attribute that provided the largest reduction in deviance (Figure
\ref{conv_class}).

Generally, the factors defining primary splits for the trees
corresponded to those with the largest reductions in deviance fitting
the regression models.

\subsubsection*{Full results}\label{full-results}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\subsection*{AIC performance for process error
structure}\label{aic-performance-for-process-error-structure}
\addcontentsline{toc}{subsection}{AIC performance for process error
structure}

\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees-1}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\subsubsection*{Full results}\label{full-results-1}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\subsection*{AIC performance for the stock-recruit
relationship}\label{aic-performance-for-the-stock-recruit-relationship}
\addcontentsline{toc}{subsection}{AIC performance for the stock-recruit
relationship}

\subsubsection*{Deviance reduction and classification
trees}\label{deviance-reduction-and-classification-trees-2}
\addcontentsline{toc}{subsubsection}{Deviance reduction and
classification trees}

\subsubsection*{Full results}\label{full-results-2}
\addcontentsline{toc}{subsubsection}{Full results}

@@

\subsection*{Bias}\label{bias-1}
\addcontentsline{toc}{subsection}{Bias}

\subsubsection*{Deviance reduction and regression
trees}\label{deviance-reduction-and-regression-trees}
\addcontentsline{toc}{subsubsection}{Deviance reduction and regression
trees}

Across all OM process error types, fishing history was the most
important single factor in reducing residual deviation in log-scale
errors for the Beverton-Holt density-independent parameter \(a\), and
also for the density-independent parameter \(b\) for R+M, R+Sel, and R+q
OMs (Table \ref{bias_SR_pars_PRD_table}). For R and R+S OMs the EM
process error assumption was equally important for estimating \(b\), but
the the reductions were small (mostly less than 4\%) for any factor and
any OM process error type. There was also no substantial reduction in
deviance with second or third order interactions of factors for either
parameter in any OM process error types.

The regression trees indicate that having contrast in fishing pressure
reduced the median error for both SR parameters for all OM types

We observed different effects of factors on deviance reduction for
errors in estimation of the two Beverton-Holt stock-recruit parameters.
When analyses were For bias of the density-dependent parameter \(b\) as
measured by the log absolute value of log-scale errors, deviance was not
reduced by any of the individual OM or EM factors except for variation
in recruitment \(\sigma_R\) for R and R+S OMs. However, second and(or)
third order interactions of the OM and EM factors also provided large
reductions in deviance for all OM process error types which indicated
that some combinations of factors could reduce bias for the \(b\)
parameter. Despite the explanation variation in the \(b\) estimation
erorrs, bias was still significant regardless of the combination of OM
and EM factors (Figures regtrees and full results).

Median natural mortality rate showed reductions in deviance for R, R+S,
and R+M OMs with second order interactions and R+M also showed
reductions with third order interactions.

\subsubsection*{Full results}\label{full-results-3}
\addcontentsline{toc}{subsubsection}{Full results}

\paragraph*{Spawning stock biomass}\label{spawning-stock-biomass}
\addcontentsline{toc}{paragraph}{Spawning stock biomass}

@@

\subsubsection*{Beverton-Holt
parameters}\label{beverton-holt-parameters}
\addcontentsline{toc}{subsubsection}{Beverton-Holt parameters}

@@

\subsubsection*{Median natural mortality
rate}\label{median-natural-mortality-rate}
\addcontentsline{toc}{subsubsection}{Median natural mortality rate}

@@

\section*{Discussion}\label{discussion}
\addcontentsline{toc}{section}{Discussion}

\subsection*{Convergence}\label{convergence-1}
\addcontentsline{toc}{subsection}{Convergence}
